---
title: "Histograms"
subtitle: "Intro to Stats, Spring 2017"
author: "Prof. Gaston Sanchez"
output: html_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Visualizing quantitative variables
> - Function `hist()`


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Intro to Descriptive Statistics

The first part of the course has to do with __Descriptive Statistics__. The 
main idea is to make a "large" or "complicated" dataset more compact and easier 
to understand by using three major tools:

- summary and frequency tables
- charts and graphics
- key numeric summaries

Describing and summarizing data can be done in many different ways. An important
consideration involves distinguishing variables between quantitative and 
qualitative ones. Depending on the type of variable, you will use certain types
of visual displays and calculate specific types of numeric values. Keep in
mind that not all graphics are valid for all classes of variables. And not all arithmetic operations will make sense on all types of variables.

In this and subsequent scripts I will cover more concepts than what the
textbook (FPP) discusses. The book focuses on:

- Histograms
- Measures of center: average and median
- Measures of spread: standard deviation

I will talk about:

- Frequency tables
- Histograms and density curves
- Measures of center: average and median
- Measures of spread: standard deviation
- Percentiles and quartiles
- The _five-number_ summary and boxplots


## More NBA Data

We are going to consider data of basketball players from the NBA. 
The data is from the season 2015-2016 and it is available in the file 
`nba_players.csv` (see the folder `data` in the course's github repository).


### Reading CSV Tables

The first step involves importing the data in R. Because the data is already 
in a tabular format (stored in a CSV file), you can use the function 
`read.csv()`. What you need is to pass the URL address of the file:

```{r}
# assembling the URL of the CSV file
# (otherwise it won't fit within the margins of this document)
repo = 'https://raw.githubusercontent.com/ucb-introstat/introstat-spring-2017/'
datafile = 'master/data/nba_players.csv'
url = paste0(repo, datafile)

# read in data set
nba = read.csv(url)
```

If you get an error message, make sure that you have the right URL
address, check the correct use of single or double quotes, and verify the
names of the functions.

_Technical Note:_ by default, `read.csv()` automatically converts any columns 
containing character strings as R factors.

The object `nba` is a data frame. Usually, after you import a data frame in R, 
you may want to use a handful of functions to inspect its properties and basic
structure:

- `str(nba)`: display overall structure of the table
- `dim(nba)`: dimensions (i.e. size) of the table
- `head(nba, n = 5)`: show first `n` rows
- `tail(nba, n = 5)`: show last `n` rows
- `names(nba)`: column names
- `colnames(nba)`: sames as `names(nba)`
- `summary(nba)`: descriptive summary of variables in `nba`

```{r}
# dimensions (num rows and columns)
dim(nba)

# names of first 10 columns
names(nba[ ,1:10])

# basic summary of first 3 columns
summary(nba[ ,1:3])
```


## Frequency tables

As we mention before, an important consideration has to do with identifying
the type of variables: quantitative -vs- qualitative. 

An example of a qualitative variable is `position`. This variable contains the
position of each player. When you inspect a qualitative variable, you typically 
start by computing a __frequency table__. A frequency table shows the frequencies 
or counts of each category. In R, we have the function `table()` to obtain
this type of table.

To obtain the frequencies of the positions simply type:

```{r}
freq_position = table(nba$position)
freq_position
```

Often, it is convenient to express the frequencies as proportions or percentages,
also referred to as __relative frequencies__.

```{r}
prop_position = freq_position / sum(freq_position)
prop_position
```

If you want to express the proportions as percentages, multiply `prop_position`
times 100:

```{r}
perc_position = 100 * prop_position
perc_position
```


## Bar-Charts and Pie-Charts

Having obtained the frequencies and/or proportions of the categories of a 
qualitative variable, we can proceed our exploration with some visual displays.
There are two most common graphics that are used to visualize frequencies:

- bar-charts
- pie-charts

To create a bar-chart in R you can use the `barplot()` function. This function
requires a numeric vector or a table of frequencies:

```{r out.width='70%', fig.align='center'}
barplot(freq_position, las = 1, border = NA, cex.names = 0.7)
```
The use of `barplot()` includes the arguments `las`, `border`, and `cex.names`:

- `las = 1`: displays the frequencies perpendicular to the y-axis
- `border = NA`: removes the black border around bars
- `cex.names = 0.7`: reduces the sizes of the category labels (so they all fit in the plot)

__Pie Chart__. The other common type of chart to see frequencies is a pie-chart. 
R provides the function `pie()` to produce these charts:

```{r out.width='35%', fig.align='center'}
pie(freq_position)
```

If you want to display the frequencies, you can do something like this:

```{r out.width='50%', fig.align='center'}
pie(freq_position, 
    labels = paste(levels(nba$position), round(perc_position, 2), "%"))
```


### Your Turn

Obtain a frequency table of the variable `team`, and plot a bar-chart of such
frequencies.


-----


## Looking at Quantitative Variables

Most of the variables in the data set `nba` are of quantitative nature. One
possibility to visually inspect those variables is to _categorize_ them and
then use a bar-chart or a pie-chart. Another possibility is to use a couple
of displays specifically dedicated to quantitative variables:

- histograms
- box-and-whisker plots (aka boxplots)


### Histograms

A [histogram](https://en.wikipedia.org/wiki/Histogram) is a type of plot that
shows the __distribution__ of numerical data.
To produce histograms, R provides the function `hist()`. The default action
of `hist()` is to plot a histogram, but you can also store its output in an R 
object. Inspecting such an object will let you see the different 
components used to plot the histogram.

Let's apply `hist()` on the column `height` (in inches) and save its output in
the object `height_hist`:

```{r out.width='70%', fig.align='center'}
height_hist = hist(nba$height, las = 1, col = 'gray80')
```

As you can tell, a histogram is very similar to a bar-chart. The common
feature is the use of bars and the use of an axis to display some sort of
frequency measure. However, a histogram is NOT a bar-chart. There are
special attributes in a histogram that makes it different from a bar-chart.

In a histogram, the bars are adjacent (no gaps between bars). Moreover,
the bars cannot be rearranged in a different order. Unlike bar-charts, what 
matters in a histogram is not the length of the bars but their areas. The
area of a bar in a histogram should be equal to the proportion of the bin.

Because we stored the output produced by `hist()` in the object `height_hist`,
we can type this object to see what output is contained in it:

```{r}
height_hist
```

- `breaks`: breaking (cutting) points of class intervals
- `counts`: number of observations in each bin
- `density`: density
- `mids`: mid-point of class intervals
- `xname`: name of object (variable) that is plotted
- `equidist`: are bins of equal width?
- `attr`: class attribute (an object of class histogram)

`hist()` produces a histogram using predefined settings. By default, it will
determine the number of bins or class intervals automatically. Like most 
histograms produced by statistical software, the default bins are
of equal size. Also, the class-intervals are right-closed (i.e. the right
endpoint is included). In the example above this means that the bins are:

- (68-70]
- (70-72]
- (72-74]
- (74-76]
- (76-78]
- (78-80]
- (80-82]
- (82-84]
- (84-86]
- (86-88]


### Steps to plot a histogram

To make a histogram, the first step is to "bin" the range of values---that is, 
divide the entire range of values into a series of class intervals---and then 
count how many values fall into each class-interval. The resulting number
of counts in a bin will give be associated to a bar in a histogram.



### FPP Textbook Histograms

I want to highlight the difference of the histograms displayed in the textbook
(FPP) with the histograms that you usually obtain with virtually all software 
programs. FPP is one of the very few introductory statistics book that shows 
histograms with bins of unequal sizes. In real life, you won't probably see 
many FPP-like histograms. In fact, I don't remember seeing a histogram of 
unequal bin-widths published in a scientific journal.

Interestingly, though, you can use `hist()` to produce FPP-style histograms in R.
To do so, you must use the `breaks` argument to provide a vector of intervals.
Let's see how to produce a histogram with the following class-intervals:

- (65-73]
- (73-75]
- (75-77]
- (77-80]
- (80-83]
- (83-90]

```{r out.width='60%', fig.align='center'}
# right-closed interval with unequal bins
height_hist2 = hist(
  x = nba$height,
  breaks = c(65, 73, 75, 77, 80, 83, 90),  # vector of class intervals
  col = 'gray80', 
  las = 1,
  axes = FALSE)
# x-axis with tick-marks that match the bins
axis(side = 1, at = c(65, 73, 75, 77, 80, 83, 90))
# y-axis
axis(side = 2, las = 1)
```

Note that we are turning off the display of the axes when using `hist()`.
In order to "manually" plot the axes, we call the function `axis()`. 
The y-axis is plotted with the argument `side = 2`. In turn, the x-axis is 
plotted with the argument `side = 1`, indicating the position of the tick 
marks `at = c(65, 73, 75, 77, 80, 83, 90)`.

Notice also that when the bins are of unequal size, `hist()` produces what
the FPP book refers to as a histogram with a __density scale__. This means that
the vertical axis does not show the frequency or count anymore. Instead, 
the height of the bars are expressed in a scale such that the area of the bars 
represent percentages.


```{r}
# bin widths
widths = height_hist2$breaks[2:7] - height_hist2$breaks[1:6]

# percentages: (bin widths) * (bar heights)
widths * height_hist2$density

# sum of percentages must equal 1
sum(widths * height_hist2$density)
```


### Histogram with right open (left-closed) intervals

An alternative histogram can be produced if we specify right-open intervals:

- [65-73)
- [73-75)
- [75-77)
- [77-80)
- [80-83)
- [83-90)

To plot a histogram with the previous bins, you must specify the argument
`right = FALSE`:

```{r out.width='60%', fig.align='center'}
# left-closed bins
left_closed = hist(
  x = nba$height,
  breaks = c(65, 73, 75, 77, 80, 83, 90),
  right = FALSE,
  col = 'gray80', 
  las = 1,
  axes = FALSE)
# x-axis
axis(side = 1, at = c(65, 73, 75, 77, 80, 83, 90))
# y-axis
axis(side = 2, las = 1)
```


### Histogram with a density scale

You can also invoke `hist()` with the default equal bins, but showing a density
scale by specifying the argument `probability = TRUE`:

```{r out.width='60%', fig.align='center'}
hist(nba$height, probability = TRUE, las = 1, col = 'gray80')
```


## Density Curves

If you take a look at the FPP book, you will see that some histograms are
sketched not with bars but with some sort of continuous curve. These figures 
are drawn as idealized or stylized histograms, just for conceptual purposes.

In practice, you can use statistical software to get similar graphics by
obtaining what is known as a _density curve_. The computational procedure
behind these type of curves involves estimating a [kernel density](https://en.wikipedia.org/wiki/Kernel_density_estimation). Estimating
kernel densities is out of the scope of the course, but I will show you how
to produce a plot that ressembles the sketched smoothed histograms of the 
textbook.

The first step is to use `density()` to obtain the kernel density estimation 
(KDE). After obtaining the KDE, you can pass it to `plot()` to get a smooth
density curve:

```{r out.width='60%', fig.align='center'}
height_density = density(nba$height)
plot(height_density, las = 1, lwd = 2)
```

Sometimes it is useful to plot a histogram, and then overlap a density
curve. This is achieved in R with the following code:

```{r out.width='60%', fig.align='center'}
hist(nba$height, probability = TRUE, las = 1, col = 'gray80')
lines(height_density, lwd = 3, col = "tomato")
```

The density curve is added via the `lines()` function. Note that we are
invoking `hist()` using the argument `probability = TRUE` in order to obtain
a ensity scale in the y-axis.




