---
title: "Histograms"
subtitle: "Intro to Stats, Spring 2017"
author: "Prof. Gaston Sanchez"
output: html_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Complete installation of R and RStudio
> - Get started with R as a scientific calculator
> - First steps using RStudio
> - Getting help in R
> - Installing packages
> - Using R script files
> - Using Rmd files
> - Get to know markdown syntax


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Histograms

R provides the function `hist()` to compute and plot histograms.

First, let's generate some artificial data. This is done by drawing a ramdon sample---with replacement---of size 100 from the numbers 1 to 10. Note the call of `set.seed()`; this allows us to obtain the same sample (for reproducibility purposes).

```{r data}
set.seed(201701)
x = sample(1:10, size = 100, replace = TRUE)
```

You can use `head()` and/or `tail()` to take a peek to the obtained numbers:

```{r}
head(x)
tail(x)
```

R provides the function `hist()` to compute and plot histograms. You can call it by passing it a vector. Here's the default behavoir of `hist()` on `x`:

```{r}
hist(x)
```

To modify the historgam:

```{r}
hist(x, las = 1, col = 'gray90')
```

To make the most out of a histogram, it is convenient to store the results in an object:

```{r}
x_hist = hist(x, las = 1, col = 'gray90', plot = FALSE)
names(x_hist)
```

You can inspect the class intervals or bins:

```{r}
x_hist$breaks
x_hist$counts
```


## Histograms like in FPP

The type of histograms discussed in the textbook are a bit special. You won't find them in other textbooks. Most statistical software programs provide default settings.

The good news is that R is very powerful, and it allows you to obtain histograms like those displayed in the book.

```{r}
histogram1 = hist(
  x, 
  breaks = seq(from = 0, to = 10, by = 2),
  las = 1,
  col = 'gray80')
```

To get a histograms with a __density scale__, you need to use the argument `probability = TRUE`:

```{r}
histogram2 = hist(
  x,
  breaks = seq(from = 0, to = 10, by = 2),
  probability = TRUE,
  las = 1, 
  col = 'gray80')
```

Let's confirm that we have a density scale in which the area of the bars add up to 1.

```{r}
sum(histogram2$density * 2)
```


-----



```{r eval = FALSE}
hist(x, probability = TRUE, las = 1)


hist(x, breaks = seq(0, 10, 2), las = 1, 
     probability = TRUE, col = 'gray80')

hist(x, breaks = seq(0, 10, 2), las = 1, 
     probability = TRUE, col = 'gray80',
     include.lowest = TRUE)

x_hist = hist(x, plot = FALSE)
plot(x_hist)

y_hist = hist(x, breaks = seq(0, 10, 2), 
              right = FALSE, plot = FALSE)
plot(y_hist)

z_hist = hist(x, breaks = seq(0, 10, 2), 
              right = TRUE, plot = FALSE)


hist(x, breaks = c(0, 2, 5, 8, 9, 10), las = 1, 
     probability = TRUE, col = 'gray80', 
     labels = 1)
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
