---
title: "Measures of Center"
subtitle: "Intro to Stats, Spring 2017"
author: "Prof. Gaston Sanchez"
output: html_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Compute the average 
> - Become familiar with the function `mean()`
> - Interpret the average as the balancing point

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As we mentioned in the previous script, the first part of the course has to do 
with __Descriptive Statistics__. The main idea is to make a "large" or 
"complicated" dataset more compact and easier to understand by using three 
major tools:

- summary and frequency tables
- charts and graphics
- key numeric summaries

In this script we will focus on various numeric summaries that are typically 
used to condense information of quantitative variables.

One common way to classify numeric summaries is in 1) measures of center, and
2) measures of spread or variability.
The idea of both types of measures is to obtain one or more numeric values that
reflect a "central" value, and the amount of "spread".

- Measures of Center
    + average or mean
    + median
- Measures of Spread
    + range
    + interquartile range
    + standard deviation (and variance)



## The Average

Perhaps the most common type of measure of center is the average or mean.
Consider a list of numbers formed by: 0, 1, 2, 3, 5, and 7. The average is
calculated as the sum of all values divided the number of values:

$$
average = \frac{0 + 1 + 2 + 3 + 5 + 7}{6} = 3
$$

You can use R to compute the previous average:

```{r}
(0 + 1 + 2 + 3 + 5 + 7) / 6
```

Algebraically, we typically denote a set of values by $x_1, x_2, \dots, x_n$, 
in which the index $n$ represents the total number of values. Using this 
notation, the formula of the average is expressed as:

$$
average = \frac{x_1 + x_2 + \dots + x_n}{n}
$$

Using summation notation, the average can be compactly expressed as:

$$
average = \sum_{i=1}^{n} \frac{x_i}{n} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$

Summation notation uses the uppercase Greek letter $\Sigma$ (sigma), 
is used as an abbreviation for the phrase "the sum of". So, in place of 
$x_1 + x_2 + \dots + x_n$, we can use summation notation as "the sum of the 
observations of the variable $x$."

In R, you can create a vector `x` to store the previous numbers:

```{r}
x = c(0, 1, 2, 3, 5, 7)
```

Then, you can use the function `sum()` to add all the values in `x`, and 
compute the average as:

```{r}
sum(x) / length(x)
```

An alternative way to compute the average in R is using the `mean()` function:

```{r}
mean(x)
```


### The Average is the balancing point

Usually, the average of a set of $n$ values $x_1, x_2, \dots, x_n$ is expressed as 
$\bar{x}$ (pronounced _x-bar_). 

To understand how the average is a type of central or mid-value, we need to 
talk about __deviations__. A deviation is the difference between an observed 
value $x_i$ and another value of reference $ref$, that is, $(x_i - ref)$.

Taking the average value $\bar{x}$ as a reference value, we can calculate the
deviations of all observations from the average: $(x_i - \bar{x})$

Given a reference value $ref$, we can also compute the sum of all deviations
around such value:

$$
\sum_{i=1}^{n} (x_i - ref)
$$

It turns out that the average is the ONLY reference value such that the sum
of deviations around it becomes zero:

$$
\sum_{i=1}^{n} (x_i - \bar{x}) = 0
$$

Let's verify that in R

```{r}
avg = mean(x)
deviations = x - avg
deviations
```

The sum of the deviations around the mean should be zero:

```{r}
sum(deviations)
```

This is the reason why we say that the average is one type of center or mid-value. 
In simpler terms, you can think of the average as the balance point of a 
distribution. The average is that point that cancels out the sum of deviations
around it.



### Your turn

We know that the average of `x` is `r mean(x)`. What happens to this average if:

- you add a constant $b$ to all values in `x`?
- you multiply the values in `x` times a constant $a$?

For instance, let's add 2 to all vaues in `x`?

```{r}
mean(x + 2)
```

Now, let's multiply by 2 all values in `x`:

```{r}
mean(x * 2)
```

Spend some time in R to examine what happens to the average of $x + k$ and 
$k \times x$ with several choices of $k$, e.g. -2, 5, 100. 

Now, let's see what happens to the average when you add a constant $b$ to all 
values in `x`, and multiply them times some constant $a$?

```{r}
mean(x)
a = 2
b = 3
mean(a*x + b)
```

Again, spend some time in R trying different values for `a` and `b`. 
What's your conclusion?


## The Median

Another common type of measure of center is the __median__. The median is the
literal middle value of an ordered distribution. By _middle value_ we mean that
half of observations are below the median, and the other half of observations
are above it.

The easiest way to calculate the median in R is with the homonym function
`median()`. Consider again the numbers in the vector `x`, the median of
this set of values is:

```{r}
x = c(0, 1, 2, 3, 5, 7)

median(x)
```


The median depends on the number of values. If you have a variable with an 
even number of values, then the median is the average of the two middle-values.
If you have a variable with an odd number of values, then the median is the
middle-value.



## More numeric summaries

Another interesting function in R that you can use to obtain descriptive
information about a variable is `summary()`. When you use this function on a
numeric vector (i.e. quantitative variable), the returned output includes:

- `Min.`: minimum
- `1st Qu.`: first quartile
- `Median`: median
- `Mean`: average
- `3rd Qu.`: third quartile
- `Max.`: maximum

```{r}
x = c(0, 1, 2, 3, 5, 7)

summary(x)
```


## Average -vs- Median

Consider a new vector `x` that contains 25 numbers: five 1's, five 2's, five 3's,
five 4's, and five 5's:

```{r}
x = rep(1:5, each = 5)
```

As you can tell, all values in `x` occur with the same frequency. And if you
get a histogram, R will plot all abrs with the same height:

```{r out.width='50%', fig.align='center'}
hist(x, breaks = c(0, 1, 2, 3, 4, 5), las = 1, col = 'gray80')
```

In this data, the average and the median are the same. In fact, this happens
all the time you have a perfect symmetric distribution:

```{r}
mean(x)
median(x)
```


Now let's add one more observation to `x` with a value of 10, and obtain the
average and the median:

```{r}
y = c(x, 10)
mean(y)
median(y)
```

Note that the average increased from `r mean(x)` to `r round(mean(y), 2)`, 
while the median remained unchanged.

Let's make it more extreme and instead of adding a value of 10 let's add a
value of 100 to `x`. The average and the median are:

```{r}
z = c(x, 100)
mean(z)
median(z)
```

Using the default plots of `hist()` in R:

```{r out.width='90%', fig.height=3, echo = FALSE, fig.align='center'}
op = par(mfrow = c(1, 3))
hist(x, las = 1, col = 'gray80')
hist(y, las = 1, col = 'gray80')
hist(z, las = 1, col = 'gray80')
par(op)
```

This is a toy example that illustrates one difference between the median and the
average. The median is more resistant (or robust) to extreme values,
but not the average. Small and large values affect the average of a distribution.

